---
title: "Databases Using R"
output: 
  html_notebook
---

## DBI & odbc packages
 
`DBI` provides a consistent set of functions that can be used to interact with a database: 

- `dbConnect()` - Use in conjuction with an R package to open a database connection
- `dbDisconnect()` - Closes a database connection

- `dbGetQuery()` - Sends a query to the database
- `dbSendQuery()`- Sends a command to the database
- `dbWriteTable()` - Creates a table
- `dbListFields()`- List fields in a table

The `odbc` allows an ODBC driver to be used to connect to a database.

Both `DBI` and `odbc` are used a connection time, the commands use are:

- `dbConnect()`
- `odbc()`



```{r}
library(DBI)
library(odbc)

con <- dbConnect(odbc(), 
                      Driver = "SQL Server", 
                      Server = "localhost\\SQLEXPRESS", 
                      Database = "datawarehouse", 
                      Trusted_Connection = "True")

con
```

```{r}
dbGetQuery(con, "Select Top 10 * From flights")
```

```{sql, connection = con}
Select top 10 * from "flights"
```

## dplyr & dbplyr

Install `dbplyr`, along with `dplyr`, to use interact with databases

```{r, eval = FALSE}
devtools::install_github("tidyverse/dbplyr")
```

```{r}
library(dplyr)
library(dbplyr)

```


```{r}
flights <- tbl(con, "vwFlights")
```

```{r}
object.size(flights)
```

```{r}
flights

```


```{r}
flights %>%
  tally()
```

```{r}
flights %>%
  group_by(name) %>%
  tally() 
```

### show_query()

```{r}
flights %>%
  group_by(name) %>%
  tally() %>%
  show_query()
```

### collect()

```{r}
top_airports <- flights %>%
  group_by(name, lon, lat) %>%
  tally() %>%
  collect()

top_airports 
```

### plotting

```{r, fig.width = 10, fig.height = 5}
library(ggplot2) 

top_airports %>%
  ggplot() +
  geom_point(aes(x = lon, y= lat, size = n, color = n), alpha = 0.5)

```

### Direct insights

```{r}
flights %>%
  group_by(name) %>%
  tally() %>%
  arrange(desc(n)) %>%  # desc() inside arrange() get the top counts
  head(10)
```

### Pipe directly into a plot

```{r}
flights %>%
  group_by(name) %>%
  tally() %>%
  arrange(desc(n)) %>%
  head(10) %>%
  collect() %>%
  ggplot() +
    geom_col(aes(x = reorder(name, n), y = n, fill = n)) +
    coord_flip()
```

### pull() command

```{r}
top_names <- flights %>%
  group_by(name) %>%
  tally() %>%
  arrange(desc(n)) %>%
  head(10) %>% 
  collect() %>%
  pull("name")

top_names
```

### Take advantage of dplyr's laziness

```{r}
top_flights <- flights %>%
  filter(name %in% top_names)

```

Use `View()` to preview some of the data in the familiar data viewer

```{r}
top_flights %>%
  head(1000) %>%
  collect() %>%
  View("top_flights")
```

Easily transition from data manipulation to plotting, without needing to import the data

```{r}
top_flights %>%
  group_by(Month) %>%
  tally() %>%
  collect() %>%
  ggplot() +
    geom_area(aes(x = Month , y = n), fill = "blue", alpha = 0.5)
  
```


Thanks to the `tidyverse` approach, an entirly new plot can be built by adding one single line, and the data is still summarized inside the database

```{r}
top_flights %>%
  filter(Cancelled == 1) %>%  # Added 
  group_by(Month) %>%
  tally() %>%
  collect() %>%
  ggplot() +
    geom_area(aes(x = Month , y = n), fill = "blue", alpha = 0.5)
  
```
## Temporary tables

```{r}
temp_flights <- flights %>%
  group_by(name) %>%
  tally() 
```

```{r}
temp_flights <- copy_to(con, temp_flights, "temp_flights")
```


```{r}
temp_flights %>%
  arrange(desc(n))
```
```{r}
dbDisconnect(con)
```





